A Novel Solution for Resonant Scattering Using Self-Consistent Boundary Conditions

### 2022-05-03 ms.tex
Adjusting fig 5 since thinner lines show the convergence at large optical depths more clearly
Incorporating Phil's most recent edits

\title{A Novel Solution to Resonant Scattering with Self-Consistent Boundary Conditions}
% the paper formerly known as Resonant Scattering in a Uniform Sphere with Large Optical Depth

@@

\begin{abstract}
%ORIGINAL: 
%The radiation transfer equation for resonant scattering is solved in spherical geometry. A monochromatic source of photons is assumed at the center of a uniform sphere of constant gas density. The intensity within the sphere and at its surface is found in the limit of large line-center optical depth using the Eddington approximation for the angular dependence. Two limits are studied: a steady source and an impulsive source. For the steady-state problem, the solution can be represented as a sum of two terms: a previously-known analytic solution of the equation with  mean intensity $J=0$ at the surface, and a novel, semi-analytic solution which enforces the correct boundary condition of zero-ingoing intensity at the surface. This solution is compared to that of the Monte Carlo method, which is valid at arbitrary optical depth. It is shown that the size of the correction is order unity when the spectral peaks approach the Doppler core and decreases slowly with line center optical depth, which may explain discrepancies in previous studies. For the impulsive problem, the time, spatial, and frequency dependence of the solution are expressed using an eigenfunction expansion in order to characterize the escape time distribution and emergent spectra of photons. It is shown that the lowest-order eigenfrequency agrees well with the decay rate found in the Monte Carlo escape time distribution at sufficiently large line-center optical depths.

\textbf{We present two novel additions to the semi-analytic solution of Lyman $\alpha$ (Ly$\alpha$) radiative transfer in spherical geometry: (1) a correction to the boundary conditions for a steady source, and (2) an eigenfunction expansion to handle the time dependence of an impulsive source.} For the steady-state problem, the solution can be represented as a sum of two terms: a previously-known analytic solution of the equation with  mean intensity $J=0$ at the surface, and a novel, semi-analytic solution which enforces the correct boundary condition of zero-ingoing intensity at the surface. This solution is compared to that of the Monte Carlo method, which is valid at arbitrary optical depth. It is shown that the size of the correction is of order unity when the spectral peaks approach the Doppler core and decreases slowly with line center optical depth, \textbf{specifically $(a \tau_0)^{-1/3}$}, which may explain discrepancies in previous studies. For the impulsive problem, the time, spatial, and frequency dependence of the solution are expressed using an eigenfunction expansion in order to characterize the escape time distribution and emergent spectra of photons. It is shown that the lowest-order eigenfrequency agrees well with the decay rate found in the Monte Carlo escape time distribution at sufficiently large line-center optical depths. \textbf{The new additions presented in this work have a broad range of important applications, since the correction term is significant any time the \lya line center optical depths exceed ${\sim}10^5$. Additionally, the characterization of the escape-time distribution highlights the potential for a Monte Carlo acceleration method, which would sample photon escape properties from distributions rather than calculating every photon scattering, reducing computational demand.}

@@

Given the abundance of hydrogen in our universe and its large cross section, \textbf{the Lyman $\alpha$ (Ly$\alpha$) line is an important component of radiation fields in wide range of astrophysical settings. \lya radiation transport is an active area of research in the study of planets, stars, galaxies, and cosmology \citep{2019SAAS...46....1D}. An example application motivating our work is the role \lya radiation transport plays in planetary atmospheres.} The outer layers of the atmosphere are central to a planet's evolution, since they can shelter the lower atmosphere from high energy radiation as well as regulate the escape of gas into space.  There are two sources of Ly$\alpha$: the star, and recombinations in the planet's atmosphere. \lya may ionize atoms and dissociate molecules, as well as exert pressure forces that drive an outflow \textbf{ \citep{2018A&A...620A.147B}}. \lya can also excite H atoms to the 2p state, creating a population of Balmer-line absorbers that can be observed via transmission spectroscopy \citep{2017ApJ...851..150H}. Due to the low gas densities in the upper atmosphere, collisional de-excitation and broadening are of secondary importance and \lya may undergo ``resonant scattering''.

@@

Due to the technical challenge of including resonant scattering, the fully three-dimensional geometry, and the presence of an outflow, numerical simulations may be required to fully understand the dynamics of these irradiated exoplanet atmospheres. The large optical depths at \lya line center impose a steep computational cost for solving radiative transfer with Monte Carlo methods directly coupled with fluid dynamics \textbf{\citep{2017MNRAS.464.2963S}}. The number of scatterings a photon undergoes is proportional to the line center optical depth, $\tau_0$, of the domain \textbf{\citep{1972ApJ...174..439A}}. Near the base of the atomic layer, the line center optical depth is ${\sim}10^8$, so most of the time is spent following photons in these cells. A method that can accurately characterize transfer through these zones without following every photon scattering has the potential to greatly accelerate the calculation.

@@

Several other works have followed \citet{1973MNRAS.162...43H}. \citet{1990ApJ...350..216N} extends the solution to media of intermediate optical depth, including the effects of scattering in the Doppler core of the line. \citet{2006ApJ...649...14D} generalize the same problem to spherical geometry, as is used here. \textbf{\cite{2020MNRAS.497.3925L} generalize both the slab and sphere solutions to arbitrary power-law density and emissivity profiles. Each of these works, and several others \citep{2020ApJS..250....9S, 2021MNRAS.504...89T}, use either the same surface boundary condition and ansatz as \citet{1973MNRAS.162...43H}, or use a solution that does not handle the frequency-dependence of the boundary condition. Our novel steady-state solution involves a frequency-dependent correction to the boundary condition that fixes an excess at the spectral peaks as compared with Monte Carlo, which is present in many of the works cited above.}

\textbf{The motivation for including time-dependence in the transfer equation is to characterize the distribution of photon escape times, which is needed to calculate the energy deposited into the gas by an escaping photon. Additionally, steady-state solutions to this problem are not always sufficient to describe all the physics of \lya transport. Time-variable, optically-thick environments necessitate a time-dependent solution to include the dynamic effects of \lya transfer. These include the optical afterglow of gamma-ray bursts \citep{2010ApJ...716..604R} and \lya sources redshifted by cosmological expansion %, which may reveal information about their neutral hydrogen halos and the intergalactic medium 
\citep{2011MNRAS.418..853X}, among others.}

@@

A better approximation to the true solution has been derived by \citet{2006ApJ...649...14D}, who generalized the closed-form solution in slab geometry found in \citet{1990ApJ...350..216N}. It satisfies a $J=0$ boundary condition at $r=R$. Again, we generalize their solution to allow emission at frequency $\sigma_{\rm s}$ away from line center. The result can be written as a sum over spatial modes,
\be \label{eq:J0_sum}
J_0 = \frac{\sqrt{6}L}{16\pi \Delta} \frac{1}{R^2}\sum_{n=1}^{\infty}n\frac{\sin{\kappa_n r}}{\kappa_n r}\exp{\left(\frac{-\kappa_n \Delta}{k}|\sigma - \sigma_s|\right)},
\ee
and
\be \label{eq:H0_sum}
H_0 = - \frac{1}{3k\phi} \frac{\partial J_0}{\partial r} = -\frac{1}{3k\phi}\frac{\sqrt{6}L}{16\pi\Delta} \frac{1}{R^2}\sum_{n=1}^{\infty}n\left(\frac{\cos{\kappa_n r}}{r} - \frac{\sin{\kappa_n r}}{\kappa_n r^2}\right)\exp{\left(\frac{-\kappa_n \Delta}{k}|\sigma - \sigma_s|\right)},
\ee
where $\kappa_n=n\pi/R$. These can be summed to give the closed form expressions
\be
J_0 & = & \frac{\sqrt{6}L}{32\pi^2 \Delta}
\frac{1}{Rr}
\left( 
\frac{ \sin(\pi r/R) }{ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_s) \right] - \cos(\pi r/R)}
\right)
\label{eq:J0}
\ee
and
\be
H_0 & = &\frac{1}{3k\phi}
\frac{\sqrt{6}L}{32\pi^2 \Delta}
\frac{1}{Rr^2}
\left( 
\frac{ \sin(\pi r/R) }{ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_s) \right] - \cos(\pi r/R)}
\right. \nonumber \\ & & \left. - \left( \frac{\pi r}{R} \right)
\frac{ \cos(\pi r/R) }{ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_s) \right] - \cos(\pi r/R)}
+ \left( \frac{\pi r}{R} \right)
\frac{ \sin^2(\pi r/R) }{ \left[ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_s) \right] - \cos(\pi r/R) \right]^2 }
\right).
\label{eq:H0}
\ee
These solutions agree with Equations (\ref{eq:Jd}) and (\ref{eq:Hd}) when the arguments of the trigonometric and hyperbolic functions are small. Again $J_0 \gg H_0$, except near $r=R$, where it goes to zero. The flux at $r=R$ can be written
\be
\nonumber
H_0(R, \sigma) & = & - \frac{1}{3k\phi}
\frac{\sqrt{6}L}{16\pi \Delta}
\frac{1}{R^3}
\sum_{n=1}^{\infty} 
n (-1)^n \exp{\left(\frac{-\kappa_n \Delta}{k}|\sigma - \sigma_s|\right)}\\
& = &  \frac{1}{3k\phi}
\frac{\sqrt{6}L}{32\pi \Delta}
\frac{1}{R^3}
\left( 
\frac{ 1 }{ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_s) \right] +1 }
\right).
\label{eq:H0surf}
\ee
\textbf{Equation (\ref{eq:H0})} will be shown to be a better approximation to the solution than Equation (\ref{eq:Hd}). It is still valid near the delta function at $r=0$, but is also a better approximation at $r=R$. $J_0$ decreases exponentially, rather than as a power-law in frequency as it does for $J_d$, giving a much smaller flux in the line wings as compared to the divergent solution.

@@

In Figure \ref{fig:sol_mc_residual_0}, the Monte Carlo spectrum is shown along with that of the solutions $H_{\rm d}$, $H_0$, and $H_0 + H_{\rm bc}$ for an optical depth of $\tau_0 = 10^7$ and photons emitted at line center $\rm x_s = 0$.  \textbf{Note that the errorbars shown on the Monte Carlo data points are $\sqrt{N}$, with $N$ being the photon count in each frequency bin, since the photons are all equally weighted.} The $H_{\rm bc}$ term is negative at the peak of the spectrum and positive in the line wing such that, when added to $H_0$, it corrects for the apparent excess of flux in the peaks of the spectrum. The solution with the correct frequency-dependent boundary condition enforced, $H_0 + H_{\rm bc}$, has lower residuals to Monte Carlo results than the other solutions, especially in the line wing. The boundary term corrects the deficit of $H_0$ in the line wings, further improving agreement with the numerical result. The residuals to the $H_0$ solution are a close match to the $H_{\rm bc}$ term, since the Monte Carlo represents the ``true'' solution, $H$, and $H_{\rm bc} = H - H_0$. It is evident that the divergent solution $H_{\rm d}$ fails in the line wings. \textbf{Also note that the ``V'' shape of the solution in the line core is due simply to the low number of points we chose to use, as the analytic solutions are not valid in this frequency regime since they utilize the damping wing approximation of the Voigt line profile. }

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{xinit.pdf}
    \caption{$P_{\rm bc}(x) = 16\pi^2R^2H_{\rm bc}(R, x)\Delta/L$ (Equation (\ref{eq:prob_spectrum}) with $H \rightarrow H_{\rm bc}$) vs. frequency $x$. \textbf{Each panel shows $\sigma_s$ further from line center, labeled by $\sigma_s=0, \tau_0, 2\tau_0$. The solution is shown at a range of line center optical depths between $\tau_0=10^5$ and $10^9$ with solid overlapping lines. The thickness of the lines decreases with increasing optical depth to more clearly show where they overlap. The location of the core-wing boundary $x_{cw}$ for each $\tau_0$ is shown with dashed vertical lines which match the color of the corresponding solution}. The axes are scaled to show that the size of the correction factor agrees with the predicted $(a\tau_0)^{-1/3}$ scaling.}
    \label{fig:xinit}
\end{figure}

@@

The emissivity for an impulsive source with energy E, source position $\vec{x}_s$, and frequency $\nu_s$ is derived in Appendix \ref{app:rteqn_derivation}. Considering a photon source at $\vec{x}_s=0$, we have (Equation \ref{eq:jem})
\be
j_{\rm em} & = & \frac{E}{4\pi} \delta^3(\vec{x}) \delta(\nu-\nu_{\rm s})\delta (t) ,
\label{eq:jem2}
\ee 
The resulting equation for $J(r,\sigma,t)$ is
\be
 \frac{-3k\phi}{c} \frac{\partial J}{\partial t} + \nabla^2 J + \left( \frac{k}{\Delta} \right)^2 \frac{\partial^2 J}{\partial \sigma^2}
& = & - \frac{\sqrt{6} kE}{4\pi \Delta^2} \delta^3(\vec{x}) \delta (\sigma - \sigma_s ) \delta (t).
\label{eq:diffusion_eqn}
\ee
We employ an expansion in terms of spherical Bessel functions in $r$ and Fourier transform in time. The zeroth spherical Bessel function is $j_0(x) = \sin{x}/x$. The expansion for $J(r, \sigma, t)$ is then
\be
\label{eq:jrsigmat_expansion}
J(r, \sigma, t) = \sum_{n=1}^{\infty} \int_{-\infty}^\infty \frac{d\omega}{2\pi} e^{-i\omega t} j_0\left(\kappa_n r\right) J(n, \sigma, \omega),
\ee
with
\be \label{eq:jnsigmaomega}
J(n, \sigma, \omega) = \frac{2\kappa_n^2}{R} \int_0^R dr\ r^2 j_0(\kappa_n r) \int_{-\infty}^\infty dt\ e^{i\omega t} J(r, \sigma, t).
\ee
Here, $\kappa_n = n\pi/R$ and $\omega$ describes the time-dependence of $J$. Though it is written as a function of the photon frequency variable $\sigma$, $J(r, \sigma, t)$ is the specific mean intensity $dE/(dA dt d\nu)$ and is a distribution in $\nu$. The Fourier coefficient $J(n, \sigma, \omega)$ has units $dE/(dA d\nu)$. Using Equation (\ref{eq:jnsigmaomega}), we obtain
\be \label{eq:diffusion_plugged_in}
 \left( \frac{3k\phi}{c}i\omega  -   \kappa_n^2 \right) J(n,\sigma,\omega)  &+& \left( \frac{k}{\Delta} \right)^2 \frac{\partial^2J(n,\sigma,\omega)}{\partial\sigma^2} = -\frac{2\kappa_n^2}{R} \frac{\sqrt{6}}{4\pi} \frac{kE}{\Delta^2} \frac{1}{4\pi} \delta(\sigma - \sigma_s).
\ee
At $\sigma=\sigma_s$, continuity must be enforced,
\be \label{eq:matching_condition_1}
J(n, \sigma^-, \omega) = J(n, \sigma^+, \omega),
\ee
and the discontinuity in $dJ/d\sigma$ due to the source is
\be \label{eq:matching_condition_2}
\frac{\partial J(n, \sigma^+, \omega)}{\partial \sigma} - \frac{\partial J(n, \sigma^-, \omega)}{\partial \sigma} & = & 
- \frac{\sqrt{6}}{8} n^2 \frac{E}{kR^3}.
\ee
At large values of $\sigma$ the line profile $\phi$ is small and Equation (\ref{eq:diffusion_plugged_in}) becomes
\be \label{eq:diffusion_at_large_sigma}
\frac{\partial^2J}{\partial\sigma^2} \approx \frac{\Delta^2\kappa_n^2}{k^2} J,
\ee
which has solutions 
\be
J(n, \sigma, \omega)\ {\sim}\ e^{\pm \kappa_n \sigma \Delta / k}.
\ee
This approximate solution implies a boundary condition at large $|\sigma|$
\be \label{eq:single_j_derivative}
\frac{\partial J}{\partial \sigma} = \mp \frac{\kappa_n\Delta}{k} J,
\ee
where a negative sign is taken for large $+\sigma$ and a positive sign is taken for large $-\sigma$ to choose the finite solution as $|\sigma|\to 0$. Numerical integrations are performed inward toward $\sigma_s$ over several domains: from large $|\sigma|$ to $\sigma_s$, from large $|\sigma|$ to 0, and from 0 to $\sigma_s$, depending on whether $\sigma_s$ is positive or negative. If $\sigma_s=0$, just two integrations are performed inward from large $|\sigma|$ to 0. Initial values for integration are obtained either by setting $J=1$ and $dJ/d\sigma$ from Equation (\ref{eq:single_j_derivative}) at large $|\sigma|$ or by matching $J$ and $dJ/d\sigma$ at 0. This gives $J$ and $J'$ on either side of $\sigma_s$, where a prime indicates the derivative $\partial/\partial \sigma$. By enforcing the matching conditions Equations (\ref{eq:matching_condition_1}) and (\ref{eq:matching_condition_2}), the eigenfunctions $J(n, \sigma, \omega)$ are obtained over the domain of photon frequencies $\sigma$. Since the solutions are linear in the starting conditions, only two integrations with different starting values are necessary.

@@

\begin{figure}
    \centering
    \includegraphics{Jsoln_n1_m100.pdf}
    \caption{Eigenfunctions $J_{nm}(x)$ for the lowest-order spatial eigenmode $n=1$, and $m=1, ..., 100$ with $x_s=0$ and $\tau_0=10^7$. \textbf{The scale of the $J_{nm}(\sigma)$ are set by the factor $E/(kR^3)$, which here is ${\sim}10^{-37}$ in units of specific mean intensity times time.}}
    \label{fig:jsoln}
\end{figure}

@@

Our characterization of the escape time distribution leads to a possible application of this work. Models of the interaction of stellar \lya with the upper atmosphere of exoplanets and the associated transmission spectrum can be constructed with a treatment of resonant scattering in spherical geometry \citep{2017ApJ...851..150H}. The Monte Carlo method can be used for this problem, but is limited by its high computational demand for large $\tau_0$ where there are many photon scatterings before escape. We seek to develop a method to accelerate the radiative transfer calculation.

>>>>

\title{A Novel Solution for Resonant Scattering Using Self-Consistent Boundary Conditions}
% the paper formerly known as Resonant Scattering in a Uniform Sphere with Large Optical Depth

@@

\begin{abstract}

\textbf{We present two novel additions to the semi-analytic solution of Lyman $\alpha$ (Ly$\alpha$) radiative transfer in spherical geometry: (1) implementation of the correct boundary condition for a steady source, and (2) solution of the time-dependent problem for an impulsive source.} For the steady-state problem, the solution can be represented as a sum of two terms: a previously-known analytic solution of the equation with  mean intensity $J=0$ at the surface, and a novel, semi-analytic solution which enforces the correct boundary condition of zero-ingoing intensity at the surface. This solution is compared to that of the Monte Carlo method, which is valid at arbitrary optical depth. It is shown that the size of the correction is of order unity when the spectral peaks approach the Doppler core and decreases slowly with line center optical depth, \textbf{specifically as $(a \tau_0)^{-1/3}$}, which may explain discrepancies seen in previous studies. For the impulsive problem, the time, spatial, and frequency dependence of the solution are expressed using an eigenfunction expansion in order to characterize the escape time distribution and emergent spectra of photons. It is shown that the lowest-order eigenfrequency agrees well with the decay rate found in the Monte Carlo escape time distribution at sufficiently large line-center optical depths. \textbf{The characterization of the escape-time distribution highlights the potential for a Monte Carlo acceleration method, which would sample photon escape properties from distributions rather than calculating every photon scattering, thereby reducing computational demand.}

@@

Given the abundance of hydrogen in the universe, \textbf{the Lyman $\alpha$ (Ly$\alpha$) line is an important component of radiation fields in a wide range of astrophysical settings. \lya radiation transport is an active area of research in the study of planets, stars, galaxies, and cosmology \citep{2019SAAS...46....1D}. An example application motivating our work is the role of \lya in planetary atmospheres.} The outer layers of the atmosphere are central to a planet's evolution, since they can shelter the lower atmosphere from high energy radiation as well as regulate the escape of gas into space.  There are two sources of Ly$\alpha$: the star, and recombinations in the planet's atmosphere. \lya may ionize atoms and dissociate molecules, as well as exert pressure forces that drive an outflow \textbf{ \citep{2018A&A...620A.147B}}. \lya can also excite H atoms to the 2p state, creating a population of Balmer-line absorbers that can be observed via transmission spectroscopy \citep{2017ApJ...851..150H}. Due to the low gas densities in the upper atmosphere, collisional de-excitation and broadening are of secondary importance and \lya may undergo ``resonant scattering''.

@@

Due to the technical challenge of including resonant scattering, the fully three-dimensional geometry, and the presence of an outflow, numerical simulations may be required to fully understand the dynamics of these irradiated exoplanet atmospheres. The large optical depths at \lya line center impose a steep computational cost for solving radiative transfer with Monte Carlo methods directly coupled with fluid dynamics \textbf{\citep{2017MNRAS.464.2963S}}. The number of scatterings a photon undergoes is proportional to the line center optical depth, $\tau_0$, of the domain \textbf{\citep{1972ApJ...174..439A}}. Near the base of the atomic layer, the line center optical depth is ${\sim}10^8$, so most of the time is spent following photons in these cells. A method that can accurately characterize transfer through these zones without following every photon scattering has the potential to greatly accelerate the calculation \textbf{(see e.g. \citealt{1968ApJ...153..783A,2002ApJ...567..922A}).}

@@

Several other works have followed \citet{1973MNRAS.162...43H}. \citet{1990ApJ...350..216N} extends the solution to media of intermediate optical depth, including the effects of scattering in the Doppler core of the line. \citet{2006ApJ...649...14D} generalize the same problem to spherical geometry, as is used here. \textbf{\cite{2020MNRAS.497.3925L} generalize both the slab and sphere solutions to arbitrary power-law density and emissivity profiles. Each of these works, and several others \citep{2020ApJS..250....9S, 2021MNRAS.504...89T}, use either the same surface boundary condition and ansatz as \citet{1973MNRAS.162...43H}, or use a solution that does not handle the frequency-dependence of the boundary condition. Our novel steady-state solution involves a frequency-dependent correction to the solution that fixes an observed excess at the spectral peaks as compared with Monte Carlo, which is present in many of the works cited above.}

\textbf{The motivation for including time-dependence in the transfer equation is to characterize the distribution of photon escape times, which is needed to calculate the radiation moments in the Monte Carlo simulation. Additionally, steady-state solutions to this problem are not always sufficient to describe all the physics of \lya transport. Time-variable, optically-thick environments necessitate a time-dependent solution to include the dynamic effects of \lya transfer. These include the optical afterglow of gamma-ray bursts \citep{2010ApJ...716..604R} and \lya sources redshifted by cosmological expansion %, which may reveal information about their neutral hydrogen halos and the intergalactic medium 
\citep{2011MNRAS.418..853X}, among others.}

@@

A better approximation to the true solution has been derived by \citet{2006ApJ...649...14D}, who generalized the closed-form solution in slab geometry found in \citet{1973MNRAS.162...43H}. It satisfies a $J=0$ boundary condition at $r=R$. Again, we generalize their solution to allow emission at frequency $\sigma_{\rm s}$ away from line center. The result can be written as a sum over spatial modes,
\be \label{eq:J0_sum}
J_0 = \frac{\sqrt{6}L}{16\pi \Delta} \frac{1}{R^2}\sum_{n=1}^{\infty}n\frac{\sin{\kappa_n r}}{\kappa_n r}\exp{\left(\frac{-\kappa_n \Delta}{k}|\sigma - \sigma_s|\right)},
\ee
and
\be \label{eq:H0_sum}
H_0 = - \frac{1}{3k\phi} \frac{\partial J_0}{\partial r} = -\frac{1}{3k\phi}\frac{\sqrt{6}L}{16\pi\Delta} \frac{1}{R^2}\sum_{n=1}^{\infty}n\left(\frac{\cos{\kappa_n r}}{r} - \frac{\sin{\kappa_n r}}{\kappa_n r^2}\right)\exp{\left(\frac{-\kappa_n \Delta}{k}|\sigma - \sigma_s|\right)},
\ee
where $\kappa_n=n\pi/R$. These can be summed to give the closed form expressions
\be
J_0 & = & \frac{\sqrt{6}L}{32\pi^2 \Delta}
\frac{1}{Rr}
\left( 
\frac{ \sin(\pi r/R) }{ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_s) \right] - \cos(\pi r/R)}
\right)
\label{eq:J0}
\ee
and
\be
H_0 & = &\frac{1}{3k\phi}
\frac{\sqrt{6}L}{32\pi^2 \Delta}
\frac{1}{Rr^2}
\left( 
\frac{ \sin(\pi r/R) }{ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_s) \right] - \cos(\pi r/R)}
\right. \nonumber \\ & & \left. - \left( \frac{\pi r}{R} \right)
\frac{ \cos(\pi r/R) }{ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_s) \right] - \cos(\pi r/R)}
+ \left( \frac{\pi r}{R} \right)
\frac{ \sin^2(\pi r/R) }{ \left[ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_s) \right] - \cos(\pi r/R) \right]^2 }
\right).
\label{eq:H0}
\ee
These solutions agree with Equations (\ref{eq:Jd}) and (\ref{eq:Hd}) when the arguments of the trigonometric and hyperbolic functions are small. Again $J_0 \gg H_0$, except near $r=R$, where it goes to zero. The flux at $r=R$ can be written
\be
\nonumber
H_0(R, \sigma) & = & - \frac{1}{3k\phi}
\frac{\sqrt{6}L}{16\pi \Delta}
\frac{1}{R^3}
\sum_{n=1}^{\infty} 
n (-1)^n \exp{\left(\frac{-\kappa_n \Delta}{k}|\sigma - \sigma_s|\right)}\\
& = &  \frac{1}{3k\phi}
\frac{\sqrt{6}L}{32\pi \Delta}
\frac{1}{R^3}
\left( 
\frac{ 1 }{ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_s) \right] +1 }
\right).
\label{eq:H0surf}
\ee
\textbf{Equation (\ref{eq:H0})} will be shown to be a better approximation to the solution than Equation (\ref{eq:Hd}). It is still valid near the delta function at $r=0$, but is also a better approximation at $r=R$. $J_0$ decreases exponentially, rather than as a power-law in frequency as it does for $J_d$, giving a much smaller flux in the line wings as compared to the divergent solution.

@@

In Figure \ref{fig:sol_mc_residual_0}, the Monte Carlo spectrum is shown along with that of the solutions $H_{\rm d}$, $H_0$, and $H_0 + H_{\rm bc}$ for an optical depth of $\tau_0 = 10^7$ and photons emitted at line center $\rm x_s = 0$.  \textbf{Note that the errorbars shown on the Monte Carlo data points are proportional to $\sqrt{N}$, with $N$ being the photon count in each frequency bin, since the photons are all equally weighted.} The $H_{\rm bc}$ term is negative at the peak of the spectrum and positive in the line wing such that, when added to $H_0$, it corrects for the apparent excess of flux in the peaks of the spectrum. The solution with the correct frequency-dependent boundary condition enforced, $H_0 + H_{\rm bc}$, has lower residuals to Monte Carlo results than the other solutions, especially in the line wing. The boundary term corrects the deficit of $H_0$ in the line wings, further improving agreement with the numerical result. The residuals to the $H_0$ solution are a close match to the $H_{\rm bc}$ term, since the Monte Carlo represents the ``true'' solution, $H$, and $H_{\rm bc} = H - H_0$. It is evident that the divergent solution $H_{\rm d}$ fails in the line wings. \textbf{Also note that the ``V'' shape of the solution in the line core is due to the low number of points plotted, as the analytic solutions are not valid in this frequency regime since they utilize the damping wing approximation of the Voigt line profile. }

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{xinit.pdf}
    \caption{$P_{\rm bc}(x) = 16\pi^2R^2H_{\rm bc}(R, x)\Delta/L$ (Equation (\ref{eq:prob_spectrum}) with $H \rightarrow H_{\rm bc}$) vs. frequency $x$. %\textbf{Each panel shows $\sigma_s$ further from line center, labeled by $\sigma_s=0, \tau_0, 2\tau_0$. The solution is shown at a range of line center optical depths between $\tau_0=10^5$ and $10^9$ with solid overlapping lines. The thickness of the lines decreases with increasing optical depth to more clearly show where they overlap. The location of the core-wing boundary $x_{cw}$ for each $\tau_0$ is shown with dashed vertical lines which match the color of the corresponding solution}. The axes are scaled to show that the size of the correction factor agrees with the predicted $(a\tau_0)^{-1/3}$ scaling.
    \textbf{Each panel shows $\sigma_s$ further from line center, labeled by $\sigma_s=0, \tau_0, 2\tau_0$. The solution is shown at a range of line center optical depths between $\tau_0=10^5$ and $10^9$ with solid lines. The location of the core-wing boundary $x_{cw}$ for each $\tau_0$ is shown with dashed vertical lines which match the color of the corresponding solution}. The axes are scaled to show that the size of the correction factor agrees with the predicted $(a\tau_0)^{-1/3}$ scaling.}
    \label{fig:xinit}
\end{figure}

@@

The emissivity for an impulsive source with energy E, source position $\vec{x}_s$, and frequency $\nu_s$ is derived in Appendix \ref{app:rteqn_derivation}. Considering a photon source at $\vec{x}_s=0$, we have (Equation \ref{eq:jem})
\be
j_{\rm em} & = & \frac{E}{4\pi} \delta^3(\vec{x}) \delta(\nu-\nu_{\rm s})\delta (t) ,
\label{eq:jem2}
\ee 
The resulting equation for $J(r,\sigma,t)$ is
\be
 \frac{-3k\phi}{c} \frac{\partial J}{\partial t} + \nabla^2 J + \left( \frac{k}{\Delta} \right)^2 \frac{\partial^2 J}{\partial \sigma^2}
& = & - \frac{\sqrt{6} kE}{4\pi \Delta^2} \delta^3(\vec{x}) \delta (\sigma - \sigma_s ) \delta (t).
\label{eq:diffusion_eqn}
\ee
We employ an expansion in terms of spherical Bessel functions in $r$ and Fourier transform in time. The zeroth spherical Bessel function is $j_0(x) = \sin{x}/x$. The expansion for $J(r, \sigma, t)$ is then
\be
\label{eq:jrsigmat_expansion}
J(r, \sigma, t) = \sum_{n=1}^{\infty} \int_{-\infty}^\infty \frac{d\omega}{2\pi} e^{-i\omega t} j_0\left(\kappa_n r\right) J(n, \sigma, \omega),
\ee
with
\be \label{eq:jnsigmaomega}
J(n, \sigma, \omega) = \frac{2\kappa_n^2}{R} \int_0^R dr\ r^2 j_0(\kappa_n r) \int_{-\infty}^\infty dt\ e^{i\omega t} J(r, \sigma, t).
\ee
Here, $\kappa_n = n\pi/R$ and $\omega$ describes the time-dependence of $J$. Though it is written as a function of the photon frequency variable $\sigma$, $J(r, \sigma, t)$ is the specific mean intensity $dE/(dA dt d\nu)$ and is a distribution in $\nu$. The Fourier coefficient $J(n, \sigma, \omega)$ has units $dE/(dA d\nu)$. Using Equation (\ref{eq:jnsigmaomega}), we obtain
\be \label{eq:diffusion_plugged_in}
 \left( \frac{3k\phi}{c}i\omega  -   \kappa_n^2 \right) J(n,\sigma,\omega)  &+& \left( \frac{k}{\Delta} \right)^2 \frac{\partial^2J(n,\sigma,\omega)}{\partial\sigma^2} = -\frac{2\kappa_n^2}{R} \frac{\sqrt{6}}{4\pi} \frac{kE}{\Delta^2} \frac{1}{4\pi} \delta(\sigma - \sigma_s).
\ee
At $\sigma=\sigma_s$, continuity must be enforced,
\be \label{eq:matching_condition_1}
J(n, \sigma^-, \omega) = J(n, \sigma^+, \omega),
\ee
and the discontinuity in $dJ/d\sigma$ due to the source is
\be \label{eq:matching_condition_2}
\frac{\partial J(n, \sigma^+, \omega)}{\partial \sigma} - \frac{\partial J(n, \sigma^-, \omega)}{\partial \sigma} & = & 
- \frac{\sqrt{6}}{8} n^2 \frac{E}{kR^3}.
\ee
At large values of $\sigma$ the line profile $\phi$ is small and Equation (\ref{eq:diffusion_plugged_in}) becomes
\be \label{eq:diffusion_at_large_sigma}
\frac{\partial^2J}{\partial\sigma^2} \approx \frac{\Delta^2\kappa_n^2}{k^2} J,
\ee
which has solutions 
\be
J(n, \sigma, \omega)\ {\sim}\ e^{\pm \kappa_n \sigma \Delta / k}.
\ee
This approximate solution implies a boundary condition at large $|\sigma|$
\be \label{eq:single_j_derivative}
\frac{\partial J}{\partial \sigma} = \mp \frac{\kappa_n\Delta}{k} J,
\ee
where a negative sign is taken for large $+\sigma$ and a positive sign is taken for large $-\sigma$ to choose the finite solution as $|\sigma|\to \infty$. Numerical integrations are performed inward toward $\sigma_s$ over several domains: from large $|\sigma|$ to $\sigma_s$, from large $|\sigma|$ to 0, and from 0 to $\sigma_s$, depending on whether $\sigma_s$ is positive or negative. If $\sigma_s=0$, just two integrations are performed inward from large $|\sigma|$ to 0. Initial values for integration are obtained either by setting $J=1$ and $dJ/d\sigma$ from Equation (\ref{eq:single_j_derivative}) at large $|\sigma|$ or by matching $J$ and $dJ/d\sigma$ at 0. This gives $J$ and $J'$ on either side of $\sigma_s$, where a prime indicates the derivative $\partial/\partial \sigma$. By enforcing the matching conditions, Equations (\ref{eq:matching_condition_1}) and (\ref{eq:matching_condition_2}), the eigenfunctions $J(n, \sigma, \omega)$ are obtained over the domain of photon frequencies $\sigma$. Since the solutions are linear in the starting conditions, only two integrations with different starting values are necessary.

@@

\begin{figure}
    \centering
    \includegraphics{Jsoln_n1_m100.pdf}
    \caption{Eigenfunctions $J_{nm}(x)$ for the lowest-order spatial eigenmode $n=1$, and $m=1, ..., 100$ with $x_s=0$ and $\tau_0=10^7$. \textbf{The scale of the $J_{nm}(\sigma)$ are set by the factor $E/(kR^3)$, which here is ${\sim}10^{-37}$ in units of ergs/(cm$^2$ Hz).}}
    \label{fig:jsoln}
\end{figure}

@@

Our characterization of the escape time distribution leads to a possible application of this work. Models of the interaction of stellar \lya with the upper atmosphere of exoplanets and the associated transmission spectrum can be constructed with a treatment of resonant scattering in spherical geometry \citep{2017ApJ...851..150H, 2021ApJ...907L..47Y}. The Monte Carlo method can be used for this problem, but is limited by its high computational demand for large $\tau_0$ where there are many photon scatterings before escape. We seek to develop a method to accelerate the radiative transfer calculation.
